An Intrinsically-Typed Solution for the\\ List-Machine Benchmark

### 2021-03-11 papers/COLA/paper.tex
Fixes based on Samuel suggestions

\begin{abstract}
Formal models are important tools in the programming language research
community. However, such models are full of intricacies and, due to that,
they are subject to subtle errors. Such failures motivated the usage of
tools to ensure the correctness of these formalisms. One way to eliminate
such errors is to encode models in a dependently-typed language in order
to ensure its ``correctness-by-construction''. In this paper, we use this
idea to build a verified interpreter for the list-machine benchmark in the
Agda programming language, comparing the results with formalizations developed
by Appel and Leroy. We formalize the 14 tasks of the benchmark
using roughly 14\% of LOC compared to a Twelf solution, and 47\% of LOC
compared to a Coq solution, even without the use of proof automation.
\end{abstract}

\begin{keyword}
Dependent types, formal semantics
\end{keyword}

@@

Using this data type\footnote{Agda supports the definition of mixfix operators.
We can use underscores to mark arguments positions.}, we can construct expressions
to denote terms that should not be considered well-typed like
\ensuremath{(\Con{Num}\;\V{1})\;\F{+}\;\Con{True}}. Using this approach, we need to specify the static semantics
as another definition, which should consider all possible cases to avoid the
definition of ill-typed terms.

@@

The list-machine, proposed by Appel et. al.~\cite{AppelDL12}, is a simple pointer virtual machine where values
are empty lists and cons-cells:
\[
a ::= nil\,\mid\,cons(a_1,a_2)
\]
Throughout this text, the meta-variable $a$ is used to denote an arbitrary value, $v$ denotes a variable
and $l$ a program label. Following common practice, all meta-variables can
appear primed or subscripted. The syntax of the virtual machine instructions are
presented next and their meaning is as usual.
\[
\begin{array}{rcll}
  \iota & ::=  & \text{jump }l                       & \text{(jump instruction)}\\
        & \mid & \text{branch-if-nil $v$~$l$}            & \text{(if $v = nil$ goto $l$)}\\
        & \mid & \text{fetch-field $v$ 0 $v'$}       & \text{(fetch the head of $v$ into $v'$)}\\
        & \mid & \text{fetch-field  $v$ 1 $v'$}      & \text{(fetch the tail of $v$ into $v'$)}\\
        & \mid & \text{cons $v_0$ $v_1$ $v'$}        & \text{(make a cons cell in v')} \\
        & \mid & \text{halt}                         & \text{(finishes execution)}\\
        & \mid & \iota_1;\iota_2                             & \text{(sequential composition)}\\
      p & ::=  & l_i \,:\,\iota\,;\,p                    & \text{(program: sequence of blocks)}\\
        & \mid & \text{end}                          & \text{(end of block list)}\\
\end{array}
\]
A program is just a sequence of blocks referenced by a unique label.

@@

Following common practice, the meta-variable $\Gamma$ denotes an
environment binding variable names to their types. Notation $\{\}$ denotes an empty environment, $v : \tau , \Gamma$
denotes the operation of including a new entry for variable $v$ with type $\tau$
in $\Gamma$ and $\Gamma [v := \tau]$ denotes the environment which is identical to $\Gamma$, except
by the entry which associates variable $v$ with type $\tau$.
Subtyping is extended to contexts as follows.
\[
\begin{array}{cc}
  \inference{}
            {\Gamma \subset_{env} \{\}}[b1]
  &
  \inference{\tau' \subset \tau & \Gamma' \subset_{env} \Gamma_2}
            {v : \tau' , \Gamma' \subset_{env} v : \tau , \Gamma_2}
            [b2]
\end{array}
\]
The variable $\Pi$ is used to denote \emph{program typings}, i.e. finite mappings between
labels and typing contexts $\Gamma$, where notation $\Pi(l) = \Gamma$ denotes that
$\Gamma$ stores the types of variables on the entry point of the block labeled by $l$.
Using the previously defined notations, the typing rules for the list-machine are defined
as a syntax directed judgment $\Pi \vdash_{\text{instr}} \Gamma \{ \iota \} \Gamma'$,
which intuitively means that the instruction $\iota$ transforms an input typing environment
$\Gamma$ into an output environment $\Gamma'$ under a fixed program typing $\Pi$. The typing
rules for the list-machine instructions are defined as follows.

The first typing rule we consider is the one for sequencing instructions inside a block. Basically,
the rule just threads the output environment from the first instruction as the input typing for the
second.
\[
\inference{\Pi \vdash_{\text{instr}} \Gamma\{ \iota_1 \}\Gamma'\:\:\:\:\:\:\:\:\Pi \vdash_{\text{instr}} \Gamma'\{ \iota_2 \}\Gamma''}
          {\Pi \vdash_{\text{instr}} \Gamma\{ \iota_1 ; \iota_2 \}\Gamma''}[seq]
\]
The type system proposed by Appel et. al.~\cite{AppelDL12} has three rules to deal with each of the possible
types assigned to the branch variable in the current typing context. The first two rules deal with the $list$ and $listcon$
types, specifying that the environment associated to the label $l$, $\Pi(l) = \Gamma_1$, is greater than
$\Gamma[v := nil]$. The third rule applies whenever $\Gamma(v) = nil$ and it also demands that $\Gamma \subset \Pi(l)$.

@@

A key feature of the list-machine type system is its subtyping (denoted by \ensuremath{\D{<:}}), which is
exposed by the least common supertype relation. We omit the Agda encoding of the subtyping relation for
space reasons, since it is just an inductive type that encodes the rules presented
in Section~\ref{sec:list}. Several lemmas about subtyping, including its decidable test,
are defined in our source-code repository~\cite{list-rep}.

@@

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{13}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathkw{data}\;\D{\_\sqcap\_\sim\_}\;\mathbin{:}\;\D{Ty}\;\rightarrow\;\D{Ty}\;\rightarrow\;\D{Ty}\;\rightarrow\;\D{Set}\;\mathkw{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Con{lub0}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\mskip1.5mu\}\;\rightarrow\;\V{t}\;\D{\sqcap}\;\V{t}\;\D{\sim}\;\V{t}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Con{lub1}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\mskip1.5mu\}\;\rightarrow\;(\Con{list}\;\V{t})\;\D{\sqcap}\;\Con{nil}\;\D{\sim}\;(\Con{list}\;\V{t}){}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Con{lub3}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\;\V{t1}\;\V{t'}\mskip1.5mu\}\;\rightarrow\;\V{t}\;\D{\sqcap}\;\V{t1}\;\D{\sim}\;\V{t'}\;\rightarrow\;{}\<[E]%
\\
\>[3]{}\hsindent{10}{}\<[13]%
\>[13]{}(\Con{list}\;\V{t})\;\D{\sqcap}\;(\Con{list}\;\V{t1})\;\D{\sim}\;(\Con{list}\;\V{t'}){}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Con{lub5}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\mskip1.5mu\}\;\rightarrow\;(\Con{listcons}\;\V{t})\;\D{\sqcap}\;\Con{nil}\;\D{\sim}\;(\Con{list}\;\V{t}){}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Con{lub7}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\;\V{t1}\;\V{t'}\mskip1.5mu\}\;\rightarrow\;\V{t}\;\D{\sqcap}\;\V{t1}\;\D{\sim}\;\V{t'}\;\rightarrow\;{}\<[E]%
\\
\>[3]{}\hsindent{10}{}\<[13]%
\>[13]{}(\Con{listcons}\;\V{t})\;\D{\sqcap}\;(\Con{listcons}\;\V{t1})\;\D{\sim}\;(\Con{listcons}\;\V{t'}){}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\orange{\texttt{-- some code omitted for brevity}}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
The presented constructors ensure that the relation is compatible with \emph{list} and \emph{listcons}
types. Also, we have a rule to ensure that the relation is commutative, as proved by a simple
Agda theorem.
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\F{\sqcap\textrm{-}comm}\;\mathbin{:}\;\V{t1}\;\D{\sqcap}\;\V{t2}\;\D{\sim}\;\V{t3}\;\rightarrow\;\V{t2}\;\D{\sqcap}\;\V{t1}\;\D{\sim}\;\V{t3}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
We omit the complete definition of \ensuremath{\F{\sqcap\textrm{-}comm}} for brevity. In the list-machine benchmark definition,
Appel et. al.~\cite{AppelDL12} also define that the least common supertype relation is sound and complete with respect to the
subtyping relation. Again, we omit the definitions of these properties, however their types are presented below.
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\F{\sqcap\textrm{-}subtype}\;\mathbin{:}\;\V{t1}\;\D{\sqcap}\;\V{t2}\;\D{\sim}\;\V{t3}\;\rightarrow\;(\V{t1}\;\D{<:}\;\V{t3})\;\D{\land}\;(\V{t2}\;\D{<:}\;\V{t3}){}\<[E]%
\\
\>[B]{}\F{\sqcap\textrm{-}least}\;\mathbin{:}\;\V{t1}\;\D{\sqcap}\;\V{t2}\;\D{\sim}\;\V{t3}\;\rightarrow\;\V{t1}\;\D{<:}\;\V{t3}\;\rightarrow\;\V{t2}\;\D{<:}\;\V{t4}\;\rightarrow\;\V{t3}\;\D{<:}\;\V{t4}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
An algorithm to construct the least supertype from two input types is
defined by function \ensuremath{\F{lub}}. Note that the \ensuremath{\F{lub}}'s type ensures that the
returned type \ensuremath{\V{t}} is indeed the least common supertype of \ensuremath{\V{t1}} and \ensuremath{\V{t2}},
thus ensuring its correctness-by-construction.

@@

For space reasons, we show how we type-check only one instruction\footnote{The curious reader can refer to our
online repository for a complete implementation.}. Function \ensuremath{\F{check\textrm{-}fetch\textrm{-}field\textrm{-}0}}
receives a program context, a typing context, and two named variables, and returns a \ensuremath{\D{TC}} value,
which is an error message or a \ensuremath{\D{CheckedInstr}} indicating that the term type-checks.

@@

\section{Extending the Machine with Indirect Jumps}\label{sec:indirect}

In this section we describe the necessary modifications to our formalization to include indirect jumps
as proposed by Appel et. al.~\cite{AppelDL12}. We start by describing the changes on the type system
and its impact on the intrinsically typed representation of programs in Section~\ref{sec:changestypes}.
Updates on the type checker and interpreter implementations are described in Sections~\ref{sec:changestypechecker}
and ~\ref{sec:changesintep}, respectively.

@@

Since our solution relies on representing the program using intrinsically typed syntax, we describe
first the changes on the machine type system. The first modification is the inclusion of a type for
program labels, which denotes continuations in a program.

@@

We omit its representation in Agda because its just a direct encoding of these previous rules
in the subtyping relation inductive type. Notice that both subtyping relations for types and
contexts are mutually inductive.

@@

\[
\begin{array}{ccc}
  \infer{\tau \sqcup \tau = \tau}{} & \infer{\top \sqcup \tau = \tau}{} & \infer{\bot \sqcup \tau = \bot}{}\\
\end{array}
\]
Next, we define the greatest lower bound relation for list types. The first rule specifies that the bottom
type is the greatest lower bound for the empty list type and non-empty list type. Two rules specifies the
compatibility of \ensuremath{\Con{list}} and \ensuremath{\Con{listcons}} type constructors with the greatest lower bound relation. The last
rule show that \ensuremath{\Con{listcons}} $\tau_3$ is the lower bound for \ensuremath{\Con{list}} $\tau_1$ and \ensuremath{\Con{listcons}} $\tau_2$.
\[
\begin{array}{cc}
  \infer{nil \sqcup (listcons\: \tau) = \bot}{} &
  \infer{(list\: \tau_1)\sqcup (list\:\tau_2) = list\:\tau_3}{\tau_1\sqcup \tau_2 = \tau_3} \\ & \\
   \infer{(listcons\: \tau_1)\sqcup (listcons\:\tau_2) = listcons\:\tau_3}{\tau_1\sqcup \tau_2 = \tau_3} &
  \infer{(list\: \tau_1)\sqcup (listcons\:\tau_2) = listcons\:\tau_3}{\tau_1\sqcup \tau_2 = \tau_3} \\ & \\
\end{array}
\]
Rules for the greatest lower bound relation for continuation types shows that bottom type is the
lower bound for list types and continuation one. The other rule for stabishes the compatibility of
the continuation type with the lower bound relation and it uses the least upper bound for typing
contexts relation.
\[
\begin{array}{ccc}
  \infer{nil\sqcup(cont\:\Gamma) = \bot}{} &
  \infer{(list\:\tau)\sqcup(cont\:\Gamma) = \bot}{} &
  \infer{(listcons\:\tau)\sqcup(cont\:\Gamma) = \bot}{} \\ & & \\
  \multicolumn{3}{c}{
    \infer{(cont\:\Gamma_1)\sqcup(cont\:\Gamma_2) = (cont\:\Gamma_3)}{\Gamma_1\cap\Gamma_2 = \Gamma_3}
  }
\end{array}
\]
We omit several rules for the greatest lower bound relation that ensure the commutativity of the
relation and that $\bot$ type is a lower bound for any type. Rules for the least upper bound for
typing contexts are simply the dual version for greatest lower bound for contexts, as shown next.
\[
  \begin{array}{cc}
    \infer{\{\} \cap \Gamma = \{\}}{} &
    \infer{\Gamma \cap \{\} = \{\}}{} \\ & \\
    \multicolumn{2}{c}{
      \infer{(\Gamma_1 , x : \tau_1) \cap (\Gamma_2 , x : t_2) = \Gamma_3, x : \tau_3}
            {\tau_1 \sqcap \tau_2 = \tau_3 & \Gamma_1 \cap \Gamma_2 = \Gamma_3}
    }
  \end{array}
\]
The Agda encoding of these relations is an immediate translation of the mathematical notation
to code and are omitted for brevity.

@@

Once we have modified the subtyping and least upper bound relations, most of the type system
rules remain unchanged from the original version. In order to support indirect jumps, Appel et. al.
approach is to modify the jump instruction and add a new instruction for loading label values into
machine registers. The typing rules for get-label and jump instructions are as follows.

\[
\begin{array}{cc}
  \infer{\Pi\vdash_{instr} \Gamma \{\textrm{get-label}\:v\:L_0\}\Gamma'}
        {\Gamma[v\,:=\,nil] = \Gamma'} &
  \infer{\Pi\vdash_{instr} \Gamma \{\textrm{get-label}\:v\:l\}\Gamma'}
        {\Pi(l) = \Gamma_1 & \Gamma[v\,:=\,cont\:\Gamma_1] = \Gamma'} \\ & \\
  \multicolumn{2}{c}{
    \infer{\Pi;\Gamma\vdash_{block}\,\textrm{jump}\:v}
          {\Gamma(v) = cont\:\Gamma_1 & \Gamma \subset \Gamma_1}
  }
\end{array}
\]
The first rule show that the type of label $L_0$, the starting label of a complete machine program,
is $nil$ and the second specifies that the type for any other label is a continuation type formed by
the context associated with label $l$ in $\Pi$. The last rule show that typing a jump instruction
requires that its register has a continuation type, $cont\:\Gamma_1$, that should be a subtype of
current block context $\Gamma$. Based on these rules, we add the following constructors to our
instruction typed syntax:

>>>>

\begin{abstract}
Formal models are important tools in the programming language research
community. However, such models are full of intricacies and, due to that,
they are subject to subtle errors. Such failures motivated the usage of
tools to ensure the correctness of these formalisms. One way to eliminate
such errors is to encode models in a dependently-typed language in order
to ensure its ``correctness-by-construction''. In this paper, we use this
idea to build a verified interpreter for the list-machine benchmark in the
Agda programming language, comparing the results with formalizations developed
by Appel and Leroy. We formalize the 14 tasks of the benchmark
using roughly 14\% of LOC compared to a Twelf solution, and 47\% of LOC
compared to a Coq solution, even without the use of proof automation. We also
describe a solution to the second version of the benchmark and compare it with
Appel's and Leroy Coq-based solution.
\end{abstract}

\begin{keyword}
Dependent types, formal semantics, Agda
\end{keyword}

@@

With this data type\footnote{Agda supports the definition of mixfix operators.
We can use underscores to mark arguments positions.}, we can construct expressions
to denote terms that should not be considered well-typed like
\ensuremath{(\Con{Num}\;\V{1})\;\F{+}\;\Con{True}}. Using this approach, we need to specify the static semantics
as another definition, which should consider all possible cases to avoid the
definition of ill-typed terms.

@@

The list-machine, proposed by Appel et. al.~\cite{AppelDL12}, is a simple pointer virtual machine where values
are empty lists and cons-cells:
\[
a ::= nil\,\mid\,cons(a_1,a_2)
\]
Throughout this text, the meta-variable $a$ is used to denote an arbitrary value, $v$ denotes a variable
and $l$ a program label. Following common practice, all meta-variables can
appear primed or subscripted. The syntax of the virtual machine instructions are
presented next and their meaning is as usual.
\[
\begin{array}{rcll}
  \iota & ::=  & \text{jump }l                       & \text{(jump instruction)}\\
        & \mid & \text{branch-if-nil $v$~$l$}            & \text{(if $v = nil$ goto $l$)}\\
        & \mid & \text{fetch-field $v$ 0 $v'$}       & \text{(fetch the head of $v$ into $v'$)}\\
        & \mid & \text{fetch-field  $v$ 1 $v'$}      & \text{(fetch the tail of $v$ into $v'$)}\\
        & \mid & \text{cons $v_0$ $v_1$ $v'$}        & \text{(make a cons cell in $v'$)} \\
        & \mid & \text{halt}                         & \text{(finishes execution)}\\
        & \mid & \iota_1;\iota_2                             & \text{(sequential composition)}\\
      p & ::=  & l_i \,:\,\iota\,;\,p                    & \text{(program: sequence of blocks)}\\
        & \mid & \text{end}                          & \text{(end of block list)}\\
\end{array}
\]
A program is just a sequence of blocks referenced by a unique label.

@@

Following the common practice, the meta-variable $\Gamma$ denotes an
environment binding variable names to their types. Notation $\{\}$ denotes an empty environment, $v : \tau , \Gamma$
denotes the operation of including a new entry for variable $v$ with type $\tau$
in $\Gamma$ and $\Gamma [v := \tau]$ denotes the environment which is identical to $\Gamma$, except
by the entry which associates variable $v$ with type $\tau$.
Subtyping is extended to contexts as follows.
\[
\begin{array}{cc}
  \inference{}
            {\Gamma \subset_{env} \{\}}[b1]
  &
  \inference{\tau' \subset \tau & \Gamma' \subset_{env} \Gamma_2}
            {v : \tau' , \Gamma' \subset_{env} v : \tau , \Gamma_2}
            [b2]
\end{array}
\]
The variable $\Pi$ is used to denote \emph{program typings}, i.e. finite mappings between
labels and typing contexts $\Gamma$, where notation $\Pi(l) = \Gamma$ denotes that
$\Gamma$ stores the types of variables on the entry point of the block labeled by $l$.
Using the previously defined notations, the typing rules for the list-machine are defined
as a syntax-directed judgment $\Pi \vdash_{\text{instr}} \Gamma \{ \iota \} \Gamma'$,
which intuitively means that the instruction $\iota$ transforms an input typing environment
$\Gamma$ into an output environment $\Gamma'$ under a fixed program typing $\Pi$. The typing
rules for the list-machine instructions are defined as follows.

The first typing rule we consider is the one for sequencing instructions inside a block. Basically,
the rule just threads the output environment from the first instruction as the input typing for the
second.
\[
\inference{\Pi \vdash_{\text{instr}} \Gamma\{ \iota_1 \}\Gamma'\:\:\:\:\:\:\:\:\Pi \vdash_{\text{instr}} \Gamma'\{ \iota_2 \}\Gamma''}
          {\Pi \vdash_{\text{instr}} \Gamma\{ \iota_1 ; \iota_2 \}\Gamma''}[seq]
\]
The type system proposed by Appel and Leroy~\cite{AppelDL12} has three rules to deal with each of the possible
types assigned to the branch variable in the current typing context. The first two rules deal with the $list$ and $listcon$
types, specifying that the environment associated to the label $l$, $\Pi(l) = \Gamma_1$, is greater than
$\Gamma[v := nil]$. The third rule applies whenever $\Gamma(v) = nil$ and it also demands that $\Gamma \subset \Pi(l)$.

@@

A key feature of the list-machine type system is its subtyping (denoted by \ensuremath{\D{<:}}), which is
exposed by the least common supertype relation. We omit the Agda encoding of the subtyping relation for
brevity, since it is just an inductive type that encodes the rules presented
in Section~\ref{sec:list}. Several lemmas about subtyping, including its decidable test,
are defined in our source-code repository~\cite{list-rep}.

@@

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{13}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathkw{data}\;\D{\_\sqcap\_\sim\_}\;\mathbin{:}\;\D{Ty}\;\rightarrow\;\D{Ty}\;\rightarrow\;\D{Ty}\;\rightarrow\;\D{Set}\;\mathkw{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Con{lub0}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\mskip1.5mu\}\;\rightarrow\;\V{t}\;\D{\sqcap}\;\V{t}\;\D{\sim}\;\V{t}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Con{lub1}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\mskip1.5mu\}\;\rightarrow\;(\Con{list}\;\V{t})\;\D{\sqcap}\;\Con{nil}\;\D{\sim}\;(\Con{list}\;\V{t}){}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Con{lub3}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\;\V{t1}\;\V{t'}\mskip1.5mu\}\;\rightarrow\;\V{t}\;\D{\sqcap}\;\V{t1}\;\D{\sim}\;\V{t'}\;\rightarrow\;{}\<[E]%
\\
\>[3]{}\hsindent{10}{}\<[13]%
\>[13]{}(\Con{list}\;\V{t})\;\D{\sqcap}\;(\Con{list}\;\V{t1})\;\D{\sim}\;(\Con{list}\;\V{t'}){}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Con{lub5}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\mskip1.5mu\}\;\rightarrow\;(\Con{listcons}\;\V{t})\;\D{\sqcap}\;\Con{nil}\;\D{\sim}\;(\Con{list}\;\V{t}){}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Con{lub7}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\;\V{t1}\;\V{t'}\mskip1.5mu\}\;\rightarrow\;\V{t}\;\D{\sqcap}\;\V{t1}\;\D{\sim}\;\V{t'}\;\rightarrow\;{}\<[E]%
\\
\>[3]{}\hsindent{10}{}\<[13]%
\>[13]{}(\Con{listcons}\;\V{t})\;\D{\sqcap}\;(\Con{listcons}\;\V{t1})\;\D{\sim}\;(\Con{listcons}\;\V{t'}){}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\orange{\texttt{-- some code omitted for brevity}}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
The presented constructors ensure that the relation is compatible with \emph{list} and \emph{listcons}
types. Also, we have a rule to ensure that the relation is commutative, as proved by a simple
Agda theorem.
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\F{\sqcap\textrm{-}comm}\;\mathbin{:}\;\V{t1}\;\D{\sqcap}\;\V{t2}\;\D{\sim}\;\V{t3}\;\rightarrow\;\V{t2}\;\D{\sqcap}\;\V{t1}\;\D{\sim}\;\V{t3}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
We omit the complete definition of \ensuremath{\F{\sqcap\textrm{-}comm}} for brevity. In the list-machine benchmark definition,
Appel and Leroy~\cite{AppelDL12} also define that the least common supertype relation is sound and complete with respect to the
subtyping relation. Again, we omit the definitions of these properties, however their types are presented below.
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\F{\sqcap\textrm{-}subtype}\;\mathbin{:}\;\V{t1}\;\D{\sqcap}\;\V{t2}\;\D{\sim}\;\V{t3}\;\rightarrow\;(\V{t1}\;\D{<:}\;\V{t3})\;\D{\land}\;(\V{t2}\;\D{<:}\;\V{t3}){}\<[E]%
\\
\>[B]{}\F{\sqcap\textrm{-}least}\;\mathbin{:}\;\V{t1}\;\D{\sqcap}\;\V{t2}\;\D{\sim}\;\V{t3}\;\rightarrow\;\V{t1}\;\D{<:}\;\V{t3}\;\rightarrow\;\V{t2}\;\D{<:}\;\V{t4}\;\rightarrow\;\V{t3}\;\D{<:}\;\V{t4}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
An algorithm to construct the least supertype from two input types is
defined by function \ensuremath{\F{lub}}. Note that the \ensuremath{\F{lub}}'s type ensures that the
returned type \ensuremath{\V{t}} is indeed the least common supertype of \ensuremath{\V{t1}} and \ensuremath{\V{t2}},
thus ensuring its correctness-by-construction.

@@

For brevity, we show how we type-check only one instruction\footnote{The curious reader can refer to our
online repository for a complete implementation.}. Function \ensuremath{\F{check\textrm{-}fetch\textrm{-}field\textrm{-}0}}
receives a program context, a typing context, and two named variables, and returns a \ensuremath{\D{TC}} value,
which is an error message or a \ensuremath{\D{CheckedInstr}} indicating that the term type-checks.

@@

\section{Extending the List-Machine with Indirect Jumps}\label{sec:indirect}

In this section we describe the necessary modifications to our formalization to include indirect jumps
as proposed by Appel and Leroy~\cite{AppelDL12}. We start by describing the changes on the type system
and its impact on the intrinsically typed representation of programs in Section~\ref{sec:changestypes}.
Updates on the type checker and the interpreter implementations are described in Sections~\ref{sec:changestypechecker}
and ~\ref{sec:changesintep}, respectively.

@@

Since our solution relies on representing the program using intrinsically typed syntax, we describe
first the changes on the list-machine type system. The first modification is the inclusion of a type for
program labels, which denotes continuations in a program.

@@

We omit their representation in Agda because they are just a direct encoding of these previous rules
in the subtyping relation inductive type. Notice that both subtyping relations for types and
contexts are mutually inductive.

@@

\[
\begin{array}{ccc}
  \infer{\tau \sqcup \tau = \tau}{} & \infer{\top \sqcup \tau = \tau}{} & \infer{\bot \sqcup \tau = \bot}{}\\
\end{array}
\]
Next, we define the greatest lower bound relation for list types. The first rule specifies that the bottom
type is the greatest lower bound for the empty list type and non-empty list type. Two rules specify the
compatibility of \ensuremath{\Con{list}} and \ensuremath{\Con{listcons}} type constructors with the greatest lower bound relation. The last
rule show that \ensuremath{\Con{listcons}} $\tau_3$ is the lower bound for \ensuremath{\Con{list}} $\tau_1$ and \ensuremath{\Con{listcons}} $\tau_2$.
\[
\begin{array}{cc}
  \infer{nil \sqcup (listcons\: \tau) = \bot}{} &
  \infer{(list\: \tau_1)\sqcup (list\:\tau_2) = list\:\tau_3}{\tau_1\sqcup \tau_2 = \tau_3} \\ & \\
   \infer{(listcons\: \tau_1)\sqcup (listcons\:\tau_2) = listcons\:\tau_3}{\tau_1\sqcup \tau_2 = \tau_3} &
  \infer{(list\: \tau_1)\sqcup (listcons\:\tau_2) = listcons\:\tau_3}{\tau_1\sqcup \tau_2 = \tau_3} \\ & \\
\end{array}
\]
Rules for the greatest lower bound relation for continuation types show that the bottom type is the
lower bound for the list and continuation types. The other rule stabishes the compatibility of
the continuation type with the lower bound relation and it uses the least upper bound for typing
contexts relation.
\[
\begin{array}{ccc}
  \infer{nil\sqcup(cont\:\Gamma) = \bot}{} &
  \infer{(list\:\tau)\sqcup(cont\:\Gamma) = \bot}{} &
  \infer{(listcons\:\tau)\sqcup(cont\:\Gamma) = \bot}{} \\ & & \\
  \multicolumn{3}{c}{
    \infer{(cont\:\Gamma_1)\sqcup(cont\:\Gamma_2) = (cont\:\Gamma_3)}{\Gamma_1\cap\Gamma_2 = \Gamma_3}
  }
\end{array}
\]
We omit several rules for the greatest lower bound relation that ensure the commutativity of the
relation and that $\bot$ type is a lower bound for any type. Rules for the least upper bound for
typing contexts are simply the dual version for greatest lower bound for contexts, as shown next.
\[
  \begin{array}{cc}
    \infer{\{\} \cap \Gamma = \{\}}{} &
    \infer{\Gamma \cap \{\} = \{\}}{} \\ & \\
    \multicolumn{2}{c}{
      \infer{(\Gamma_1 , x : \tau_1) \cap (\Gamma_2 , x : t_2) = \Gamma_3, x : \tau_3}
            {\tau_1 \sqcap \tau_2 = \tau_3 & \Gamma_1 \cap \Gamma_2 = \Gamma_3}
    }
  \end{array}
\]
The Agda encoding of these relations is an immediate translation of the mathematical notation
to code and are omitted for brevity.

@@

Once we have only modified the subtyping and the least upper bound relations, most of the type system
rules remained unchanged from the original version. In order to support indirect jumps, Appel and Leroy.
approach is to modify the jump instruction and add a new instruction for loading label values into
machine registers. The typing rules for get-label and jump instructions are as follows.

\[
\begin{array}{cc}
  \infer{\Pi\vdash_{instr} \Gamma \{\textrm{get-label}\:v\:L_0\}\Gamma'}
        {\Gamma[v\,:=\,nil] = \Gamma'} &
  \infer{\Pi\vdash_{instr} \Gamma \{\textrm{get-label}\:v\:l\}\Gamma'}
        {\Pi(l) = \Gamma_1 & \Gamma[v\,:=\,cont\:\Gamma_1] = \Gamma'} \\ & \\
  \multicolumn{2}{c}{
    \infer{\Pi;\Gamma\vdash_{block}\,\textrm{jump}\:v}
          {\Gamma(v) = cont\:\Gamma_1 & \Gamma \subset \Gamma_1}
  }
\end{array}
\]
The first rule shows that the type of label $L_0$, the starting label of a complete machine program,
is $nil$ and the second specifies that the type for any other label is a continuation type formed by
the context associated with label $l$ in $\Pi$. The last rule show that typing a jump instruction
requires that its register has a continuation type, $cont\:\Gamma_1$, that should be a subtype of
the current block context $\Gamma$. Based on these rules, we add the following constructors to our
instruction typed syntax:
